Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2009-12-08 ltx/BrainIDE.tex
added multilayer model to latex doco

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
f\left( {{v_t}\left( {r'} \right)} \right) = \frac{{{\nu _0}}}{{1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - {v_t}\left( {r'} \right)} \right)} \right)}}.
\end{equation}
%\begin{equation}\label{RateOfPreSynPops}
%p\left( {v\left( {r,\bar t} \right)} \right) = {\nu_0}u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
%\begin{equation}\label{HeavisideDef}
%u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
%   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
%   0 & {otherwise}  \\
%\end{array}} \right.
Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {r,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  +e\left( {r,t} \right).
\end{equation}

@@

\section{State-Space Model}
We begin the derivation of the state-space model with the discrete time, continuous space IDE neural field model
\begin{equation}\label{SystemEq}
{v_{t + 1}}\left( {\bf{r}} \right) = \int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}}  - \alpha {v_t}\left( {\bf{r}} \right) + e_t\left( {\mathbf{r}} \right),
\end{equation}
where $e_t(r)\sim \mathcal{GP}(0,R)$. The observation equation for the model is
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )v_t( \mathbf{r}\rm ) d\mathbf{r}  + \varepsilon ( k ),
\end{equation}
where $\varepsilon( k )\sim\mathcal{N}(0,\Sigma)$, $k$ indexes the iEEG electrodes, and $m(k,r)$ is the observation kernel relating the neural field to EEG measurements accounting for electrodes characteristics.
To infer the state from observations we can write the likelihood function
\begin{equation}\label{Likelihood}
p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
\end{equation}
where
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
\end{eqnarray}
We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can ellaborate), so we need to decompose the field into weighted basis functions.

>>>>

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
f\left( {{v_t}\left( {r'} \right)} \right) = \frac{{{\nu _0}}}{{1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - {v_t}\left( {r'} \right)} \right)} \right)}}.
\end{equation}
%\begin{equation}\label{RateOfPreSynPops}
%p\left( {v\left( {r,\bar t} \right)} \right) = {\nu_0}u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
%\end{equation}
%\begin{equation}\label{HeavisideDef}
%u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
%   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
%   0 & {otherwise}  \\
%\end{array}} \right.
%\end{equation}
Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {r,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  +e\left( {r,t} \right).
\end{equation}
\subsection{Multi-Layer Model}
\begin{equation}\label{ExcitatoryLayer}
\frac{dv_e\left( {r,t} \right)}{{dt}} + \alpha_ev_e\left( {r,t} \right) = \int_\Omega  {w_{ee}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ei}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_e\left( {r,t} \right)
\end{equation}
\begin{equation}\label{InhibitoryLayer}
\frac{dv_i\left( {r,t} \right)}{{dt}} + \alpha_iv_i\left( {r,t} \right) = \int_\Omega  {w_{ie}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ii}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_i\left( {r,t} \right)
\end{equation}
The subscript $ei$ denotes connectivity from an inhibitory neural element to an excitatory neural element. We can write the model in matrix form
\begin{equation}\label{matrixEquation}
\left[ {\begin{array}{*{20}{c}}
   {\frac{{d{v_e}\left( t \right)}}
{{dt}}}  \\
   {\frac{{d{v_i}\left( t \right)}}
{{dt}}}  \\

 \end{array} } \right] = \int_\Omega  {\left[ {\begin{array}{*{20}{c}}
   {{w_{ee}}\left( {r,r'} \right)} & {{w_{ei}}\left( {r,r'} \right)}  \\
   {{w_{ie}}\left( {r,r'} \right)} & {{w_{ii}}\left( {r,r'} \right)}  \\

 \end{array} } \right]} \left[ {\begin{array}{*{20}{c}}
   {f\left( {{v_e}\left( {r',t} \right)} \right)}  \\
   {f\left( {{v_i}\left( {r',t} \right)} \right)}  \\

 \end{array} } \right]dr' - \left[ {\begin{array}{*{20}{c}}
   {{\alpha _e}{v_e}\left( {r,t} \right)}  \\
   {{\alpha _i}{v_i}\left( {r,t} \right)}  \\

 \end{array} } \right] + \left[ {\begin{array}{*{20}{c}}
   {{\varepsilon _e}\left( t \right)}  \\
   {{\varepsilon _i}\left( t \right)}  \\

 \end{array} } \right]
\end{equation}
The iEEG observation equation is
\begin{equation}\label{iEEGObservationEquation}
    y(k,t) = \int_{\Omega} m(k,r)\left(v_e(r,t) + v_i(r,t)\right)dr' + w(k,t) 
\end{equation}

@@

\section{State-Space Model}
We begin the derivation of the state-space model with the discrete time, continuous space IDE neural field model. A one-step Euler method was used to transform the ODE to a discrete time form.
\begin{equation}\label{SystemEq}
\frac{v_{t + 1}\left( \mathbf{r} \right) - v_t(\mathbf{r})}{T_s} = \int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}}  - \alpha {v_t}\left( {\bf{r}} \right) + e_t\left( {\mathbf{r}} \right),
\end{equation}
\begin{equation}\label{SystemEq}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}} - T_s\alpha {v_t}\left( {\bf{r}} \right) + v_t(\mathbf{r}) + T_se_t\left( {\mathbf{r}} \right),
\end{equation}
\begin{equation}\label{SystemEq}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}} + (1 - T_s\alpha) v_t\left( {\bf{r}} \right) + T_se_t\left( {\mathbf{r}} \right),
\end{equation}
\begin{equation}\label{SystemEq}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}} + \lambda v_t\left( {\bf{r}} \right) + T_se_t\left( {\mathbf{r}} \right).
\end{equation}
where $e_t(r)\sim \mathcal{GP}(0,R)$, $\lambda = 1-T_s\alpha$ and $T_s$ is the time step or sampling step.

The observation equation for the model is
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )v_t( \mathbf{r}\rm ) d\mathbf{r}  + \varepsilon ( k ),
\end{equation}
where $\varepsilon( k )\sim\mathcal{N}(0,\Sigma)$, $k$ indexes the iEEG electrodes, and $m(k,r)$ is the observation kernel relating the neural field to EEG measurements accounting for electrodes characteristics.

To infer the state from observations we can write the likelihood function
\begin{equation}\label{Likelihood}
p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
\end{equation}
where
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
\end{eqnarray}
We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can elaborate), so we need to decompose the field into weighted basis functions.
